---
title: "eika_can_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eika_can_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Background
Facing climate change and nature loss, reflecting on the interlinked climate-nature crisis, calls actions in different economic sectors. To enhance corporate transparency, investor confidence and finally more sustainable practices across different sectors, the European Sustainability Reporting Standards (ESRS) under the Corporate Sustainability Reporting Directive (CSRD) have been established. The standard requires estimation and reporting regarding the companies effect on environment and society.

The following application (CAN tool), at the proof of concept (POC) stage, facilitates reporting and tracking impacts of building activities on climate and nature. Thus, climate (E1) and nature risks (E4) of the ESRS are partially covered with the tool.

## Novelty
The tool uses spatial explicit data of the area of building activity (plot or region of impact) and locally valid climate and nature data. Consequently, the reporting of climate and nature impact summarizes project-based impact estimates.


## Goals
* Screen projects for potential high risk exposure/vulnerabilities regarding climate change
* Screen projects for potential negative impacts on valuable nature
* Facilitate reporting of a whole project portfolio

# Data
To calculate potential risk exposure of climate change and impacts on valuable nature of specific projects, local valid and high resolution data layers are used. However, in the pilot phase of this project, data is only available for Åfjord, Lillestrøm and Verdal. 

## General data
This data includes information about the target municipalities and the plots to locate projects
```{r, echo=FALSE}
library(knitr)

data1 <- data.frame(
  Data = c("Matrikkelen - Eiendomskart Teig", "Administrative enheter kommuner"),
  Data_owner = c("Norwegian Mapping Authority", "Norwegian Mapping Authority"),
  Source = c("https://kartkatalog.geonorge.no/metadata/matrikkelen-eiendomskart-teig/74340c24-1c8a-4454-b813-bfe498e80f16", "https://kartkatalog.geonorge.no/metadata/administrative-enheter-kommuner/041f1e6e-bdbc-4091-b48f-8a5990f3cc5b?search=kommun"),
  Version = c("31.03.2025","31.03.2025")
)

kable(data1, format = "markdown")
```


## Climate
```{r, echo=FALSE}
library(knitr)

data2 <- data.frame(
  Data = c("Flomsoner", "Kvikkleire"),
  Data_owner = c("Norwegian Water Resources and Energy Directorate", "Norwegian Water Resources and Energy Directorate"),
  Source = c("https://kartkatalog.geonorge.no/metadata/flomsoner/e95008fc-0945-4d66-8bc9-e50ab3f50401?search=flomson", "https://kartkatalog.geonorge.no/metadata/kvikkleire/a29b905c-6aaa-4283-ae2c-d167624c08a8?search=kvikk"),
  Version = c("28.3.2025","28.3.2025")
)

kable(data2, format = "markdown")
```

## Nature
```{r, echo=FALSE}
library(knitr)

data3 <- data.frame(
  Data = c("Naturvernområder", "Friluftslivsområder - kartlagte", "Inngrepsfri natur i Norge","Villreinområder","Naturtyper - verdsatte","Statlige planretningslinjer for differensiert forvaltning av strandsonen langs sjøen","Hovedøkosystemer i Norge","Naturskog","Myr","Vassdragsnatur","Rødlistearter"),
  Data_owner = c("Norwegian Environment Agency","Norwegian Environment Agency","Norwegian Environment Agency","Norwegian Environment Agency","Norwegian Environment Agency","Ministry of Local Government and Regional Development","Norwegian Environment Agency","Norwegian Environment Agency","Norwegian Institute for Nature Research","Norwegian Institute for Nature Research", "Artsdatabanken"),
  Source = c("https://kartkatalog.geonorge.no/metadata/naturvernomraader/5857ec0a-8d2c-4cd8-baa2-0dc54ae213b4", "https://kartkatalog.geonorge.no/metadata/friluftslivsomraader-kartlagte/91e31bb7-356f-4478-bcba-d5c2de6e91bc?search=kartlag", "https://kartkatalog.geonorge.no/metadata/inngrepsfri-natur-i-norge/277bda73-b924-4a0e-b299-ea5441de2d3b?search=inngr","https://kartkatalog.geonorge.no/metadata/villreinomraader/fc59e9a4-59df-4eb3-978a-1c173b84bf4e?search=villrei","https://kartkatalog.geonorge.no/metadata/naturtyper-ku-verdi-wms/aea3702c-3072-4a00-88bc-07cc3992b5a4?search=ku%20natur","https://kartkatalog.geonorge.no/metadata/statlige-planretningslinjer-for-differensiert-forvaltning-av-strandsonen-langs-sjoeen/f50f228a-9482-4821-97df-41166c1f5a9b?search=strands","https://kartkatalog.geonorge.no/metadata/hovedoekosystemer-i-norge-versjon-1/ce4261f9-486d-4f28-8687-d6d0f277f667?search=%C3%B8kos","https://kartkatalog.geonorge.no/metadata/naturskog-v1/a0062ac4-8ee0-408f-9373-4c8b8c3088d8?search=natursko","no online resource","no online resource","https://kartkatalog.geonorge.no/metadata/roedlistearter/00b0ab63-de0b-4f12-ac40-655de5163b8d?search=r%C3%B8d"),
  Version = c("01.01.2021", "01.01.2021", "01.01.2023","01.01.2021","21.04.2022","11.03.2025","14.12.2023","10.01.2025","","01.01.2025","16.03.2025")
)

kable(data3, format = "markdown")
```

The data was partitioned by the three case study municipalities (Åfjord, Lillestrøm and Verdal)


# Calculations
The calculations of climate and nature risk follows a three step approach:

1) Selection of a municipality
Load the local data and visualize the data through WMS layers

2) Select a plot from the "teig" data using brusknummer, gårdsnummer and teignummer
* Load the plot geometry
* Calculate min distance of the plot to each data layer (climate & nature)

```{r, eval=FALSE}
distance <- st_distance(plot_geom, layers[[i]]) %>% min()  # Closest distance
```

* if min dist == 0 calculate intersection polygon and intersection polygon area
```{r, eval=FALSE}
 if(as.numeric(distance) == 0){
        # Compute intersection AREA
        intersection <- TRUE
        intersect_poly<-st_intersection(geom, layers[[i]])%>%select()%>%mutate(layer_id = names(layers[i])) #geometry
        intersection_area <- st_area(intersect_poly) #area


      }else{
        intersection <- FALSE
        intersect_poly<- NULL
        intersection_area <- NA

      }
```

* Calculate area per land-use/land-cover (raster) class within the plot
```{r, eval=FALSE}
overlap_stats <- exactextractr::exact_extract(raster, plot_geom)
```


3) Display results

# Outputs
The outputs show red and green "flags" according to intersection and non-intersection of the project area with one or several climate or nature layers. For each layer the intersection, intersection area and areas per land-cover/land-use class can be saved as CSV. In addition the results for each plot are saved on a DUCK data base (duckdb)

##  Covered ESRS data points
The tools covers the following data points from the ESRS reporting standard on numerical, spatially explicit basis.

```{r, echo = FALSE}
library(knitr)

data4 <- data.frame(
  ID = c("E1.IRO-1_02 ", "E1.IRO-1_04","E4.SBM-3_01","E4.SBM-3_02","E4.SBM-3_03","E4.SBM-3_04","E4.SBM-3_05","E4.SBM-3_06","E4.IRO-1_14","E4-5_01","E4-5_02"),
  Description = c("Description of process in relation to climate-related physical risks in own operations and along value chain ", "","Localisation of a financing project ","Type definition of a financing project","Due diligence of projects nature risk","Loss of valuable natural areas","Conversion of natural surfaces into sealed surfaces","Potential negative impact on red-listed species / valuable nature areas", "The project is within valuable nature areas","Distance to nature protected areas and valuable nature areas","Project area that lies within nature protected or valuable nature areas"),
  Data_used = c("Temperature and flood data ", "","Municipal and plot geometries","User input","Valuable nature data","Valuable nature data","Valuable nature data","Land-use / land cover data","Potential negative impact on red-listed species / valuable nature areas","Nature protected areas","Nature protected areas"),
  Calculations_used = c("","","Selection","Selection","Distance, intersection and area","Distance, intersection and area","Distance, intersection and area","Intersection","Distance, intersection and area","Distance, intersection and area","Distance, intersection and area")
)

kable(data4, format = "markdown")

```

